Exploring the Utility of Machine Learning Across Varied Data Formats

### 2024-05-26 homework2/paper/paper.tex
finish

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}
\pagebreak
The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets. By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

@@

@@

\section{Evaluation}

% ------------------------------------------------------------------------------
\section{Conclusions}

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://curs.upb.ro/2023/course/view.php?id=13749}{Moodle - Artifical Intelligence Course}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Linear_regression}{Wikipedia - Linear Regression}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Linear_classifier}{Wikipedia - Linear Classifier}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Ridge_regression}{Wikipedia - Ridge Regression}
    \bibitem{}
    \href{https://www.semanticscholar.org/reader/2c9022fe0af15568a885e59d475ec8f95726e51b}{Metrics for Multi-Class Classification: An Overview}
    \bibitem{}
    \href{https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html}{Pandas - DataFrame}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Correlation}{Wikipedia - Correlation}
    \bibitem{}
    \href{https://thinkingneuron.com/how-to-measure-the-correlation-between-two-categorical-variables-in-python/}{Thinking Neuron - How to Measure the Correlation Between Two Categorical Variables in Python}
    \bibitem{}
    \href{https://www.stratascratch.com/blog/chi-square-test-in-python-a-technical-guide/}{StrataScratch - Chi-Square Test in Python: A Technical Guide}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%27s\_V}{Wikipedia - Cramér's V}
    \bibitem{}
    \href{https://saturncloud.io/blog/how-to-detect-and-exclude-outliers-in-a-pandas-dataframe/}{Saturn Cloud - How to Detect and Exclude Outliers in a Pandas DataFrame}
    \bibitem{}
    \href{https://www.kaggle.com/discussions/questions-and-answers/159183}{Kaggle - When to standardize test and train data?}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/preprocessing.html#standardization-or-mean-removal-and-variance-scaling}{Scikit-Learn - Standardization or Mean Removal and Variance Scaling}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.preprocessing.OneHotEncoder.html}{Scikit-Learn - OneHotEncoder}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.preprocessing.LabelEncoder.html}{Scikit-Learn - LabelEncoder}
    \bibitem{}
    \href{https://towardsdatascience.com/understanding-confusion-matrix-a9ad42dcfd62}{Towards Data Science - Understanding Confusion Matrix}

>>>>

\pagebreak

@@

The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets. By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}

@@

\section{Evaluation}
The evaluation of machine learning models is a critical step in assessing their
performance and effectiveness. By comparing the model's predictions to the actual
ground truth, we can determine the model's accuracy, precision, recall, and other
metrics that quantify its performance. This section delves into the evaluation
of the Logistic Regression and Multi-Layered Perceptron (MLP) models on the
Salary Prediction and Stroke Prediction datasets.

\subsection{Hyperparameter Tuning}
Hyperparameters are parameters that are set before the learning process begins.
They control the learning process and the behavior of the model. Hyperparameter
tuning is the process of selecting the optimal hyperparameters for a machine
learning model to achieve the best performance. This process involves searching
through different hyperparameter configurations and evaluating the model's
performance on a validation set to find the optimal settings.

In the context of the Logistic Regression (manual implementation), the 
hyperparameters that were used are:

\begin{itemize}
    \item \textit{learning\_rate:} The rate at which the model updates the weights - 0.01
    \item \textit{num\_iterations} The number of iterations the model trains for - 10000
    \item \textit{regularization} The regularization parameter to prevent overfitting - 0.1
\end{itemize}

In the context of the Logistic Regression (Scikit-Learn implementation), the majority of
hyperparameters that were used are the default values provided by the Scikit-Learn
library. The only hyperparameters that were set are:

\begin{itemize}
    \item \textit{solver:} The optimization algorithm used in the model - 
    'saga' for SalaryPrediction and 'sag' for AVC
    \item \textit{max\_iter:} The maximum number of iterations for the optimization algorithm - 200 for SalaryPrediction and 500 for AVC
    \item \textit{C:} The regularization parameter to prevent overfitting - 0.4342470001 for SalaryPrediction and  2.56050926 for AVC
\end{itemize}

In the context of the Multi-Layered Perceptron (manual implementation), the
hyperparameters that were used are:

\begin{itemize}
    \item \textit{hidden\_sizes:} The sizes of the hidden layers are defined as a list - [256, 128, 64]
    \item \textit{num\_epochs:} The number of epochs the model trains for - 100
    \item \textit{learning\_rate:} The rate at which the model updates the weights - 0.01
    \item \textit{Loss Function	:} The loss function used to optimize the model - Negative Log-Likelihood
\end{itemize}

In the context of the Multi-Layered Perceptron (Scikit-Learn implementation), the 
hyperparameters that were used are the default values provided by the Scikit-Learn
library.

\subsection{Confusion Matrix}
A confusion matrix is a table that summarizes the performance of a classification
model on a set of test data for which the true values are known. It provides
insights into the model's predictions, including true positive, true negative,
false positive, and false negative instances. These metrics are essential for
evaluating the model's performance and identifying potential areas for improvement.

In the context of the Logistic Regression and Multi-Layered Perceptron models, we
generated confusion matrices to analyze the model's predictions on the test data.
The confusion matrices provide a detailed breakdown of the model's performance,
highlighting the number of correct and incorrect predictions for each class.

In Figures \ref{fig:confusion_matrix_logistic_manual_salary} and 
\ref{fig:confusion_matrix_logistic_scikit_salary} we can see the confusion matrices
for the Logistic Regression model on the Salary Prediction dataset, implemented
manually and with Scikit-Learn, respectively. In Figures 
\ref{fig:confusion_matrix_mlp_manual_salary} and \ref{fig:confusion_matrix_mlp_scikit_salary}
we can see the confusion matrices for the Multi-Layered Perceptron model on the
Salary Prediction dataset, implemented manually and with Scikit-Learn, respectively.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_SalaryPrediction/LogisticRegression/confusion_matrix_manual.png}
    \caption{Confusion Matrix for the Logistic Regression model on the Salary Prediction dataset (Manual Implementation)}
    \label{fig:confusion_matrix_logistic_manual_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_SalaryPrediction/LogisticRegression/confusion_matrix_scikit-learn.png}
    \caption{Confusion Matrix for the Logistic Regression model on the Salary Prediction dataset (Scikit-Learn Implementation)}
    \label{fig:confusion_matrix_logistic_scikit_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_SalaryPrediction/MLP/confusion_matrix_manual.png}
    \caption{Confusion Matrix for the Multi-Layered Perceptron model on the Salary Prediction dataset (Manual Implementation)}
    \label{fig:confusion_matrix_mlp_manual_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_SalaryPrediction/MLP/confusion_matrix_scikit-learn.png}
    \caption{Confusion Matrix for the Multi-Layered Perceptron model on the Salary Prediction dataset (Scikit-Learn Implementation)}
    \label{fig:confusion_matrix_mlp_scikit_salary}
\end{figure}

We can see that the Logistic Regression model implemented with Scikit-Learn achieved
almost the same results as the manual implementation and the MLP model implemented
with Scikit-Learn. All 4 models achieved a high number of true positive predictions
for the negative class, indicating that the models are effective at identifying
negative instances in the dataset. The true positive predictions for the positive
class are lower, but the models still achieved a reasonable prediction but the
MLP model with manual implementation achieved the worst results. It may be due to
the hyperparameters that were set for the model. The false positive and false negative
predictions are also lower than the true positive and true negative predictions, which
is a good sign for the model's performance.

In Figures \ref{fig:confusion_matrix_logistic_manual_stroke} and
\ref{fig:confusion_matrix_logistic_scikit_stroke} we can see the confusion matrices
for the Logistic Regression model on the Stroke Prediction dataset, implemented
manually and with Scikit-Learn, respectively. In Figures
\ref{fig:confusion_matrix_mlp_manual_stroke} and \ref{fig:confusion_matrix_mlp_scikit_stroke}
we can see the confusion matrices for the Multi-Layered Perceptron model on the
Stroke Prediction dataset, implemented manually and with Scikit-Learn, respectively.

Because of the high class imbalance in the Stroke Prediction dataset, the models
can not predict the positive class effectively (the positive class is less than 5\%
of the dataset). On the other hand, the models can predict the negative class
effectively.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_AVC/LogisticRegression/confusion_matrix_manual.png}
    \caption{Confusion Matrix for the Logistic Regression model on the Stroke Prediction dataset (Manual Implementation)}
    \label{fig:confusion_matrix_logistic_manual_stroke}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_AVC/LogisticRegression/confusion_matrix_scikit-learn.png}
    \caption{Confusion Matrix for the Logistic Regression model on the Stroke Prediction dataset (Scikit-Learn Implementation)}
    \label{fig:confusion_matrix_logistic_scikit_stroke}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_AVC/MLP/confusion_matrix_manual.png}
    \caption{Confusion Matrix for the Multi-Layered Perceptron model on the Stroke Prediction dataset (Manual Implementation)}
    \label{fig:confusion_matrix_mlp_manual_stroke}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../tema2_AVC/MLP/confusion_matrix_scikit-learn.png}
    \caption{Confusion Matrix for the Multi-Layered Perceptron model on the Stroke Prediction dataset (Scikit-Learn Implementation)}
    \label{fig:confusion_matrix_mlp_scikit_stroke}
\end{figure}

\subsection{Evaluation Metrics}
To evaluate the performance of the machine learning models, we employ a range of
evaluation metrics that provide insights into different aspects of the model's
performance. These metrics include:

\begin{itemize}
    \item \textit{Accuracy:} The proportion of correctly classified instances
    out of the total instances. It provides a general measure of the model's
    correctness.
    \item \textit{Precision:} The proportion of true positive predictions out
    of all positive predictions. It measures the model's ability to avoid false
    positives.
    \item \textit{Recall:} The proportion of true positive predictions out of
    all actual positive instances. It measures the model's ability to capture
    all positive instances.
    \item \textit{F1-Score:} The harmonic mean of precision and recall. It provides
    a balanced measure of the model's performance.
\end{itemize}

These evaluation metrics help us understand the strengths and weaknesses of the
machine learning models and guide us in improving their performance. By analyzing
these metrics, we can identify areas for optimization and fine-tuning to enhance
the model's predictive capabilities.

In the following tables, Table \ref{tab:evaluation_metrics_salary} and Table
\ref{tab:evaluation_metrics_stroke}, we present the evaluation metrics for the 
Logistic Regression
and Multi-Layered Perceptron models on both prediction tasks: Salary Prediction and
Stroke Prediction.

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{4cm}||>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|}
        \hline
        \multicolumn{5}{|c|}{Evaluation Metrics for Salary Prediction Dataset - Train set} \\
        \hline
        Model & Accuracy & Precision & Recall & F1-Score \\
        \hline\hline
        Logistic Regression (Manual) & 0.829353669 & 0.700867052 & 0.504945340 & 0.586989409 \\
        \hline
        Logistic Regression (Scikit-Learn) & 0.8371046 & 0.707940780 & 0.54763144 & 0.617552098 \\
        \hline
        Multi-Layered Perceptron (Manual) & 0.759844980 & \textbf{1.0} & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Scikit-Learn) & \textbf{0.91098887} & 0.85537918 & \textbf{0.75741801} & \textbf{0.80342352} \\
        \hline
        \hline
        \multicolumn{5}{|c|}{Evaluation Metrics for Salary Prediction Dataset - Test set} \\
        \hline
        Model & Accuracy & Precision & Recall & F1-Score \\
        \hline\hline
        Logistic Regression (Manual) & \textbf{0.831} & 0.712250712 & 0.513347022 & 0.596658711 \\
        \hline
        Logistic Regression (Scikit-Learn) & 0.825 & 0.676092544 & \textbf{0.54004106} & \textbf{0.60045662} \\
        \hline
        Multi-Layered Perceptron (Manual) & 0.7565 & \textbf{1.0} & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Scikit-Learn) & 0.811 & 0.635235732 & 0.525667351 & 0.575280898 \\
        \hline
        \caption{Evaluation Metrics for Logistic Regression and Multi-Layered Perceptron Models on Salary Prediction Dataset}
        \label{tab:evaluation_metrics_salary} \\
    \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{4cm}||>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{2cm}|}
        \hline
        \multicolumn{5}{|c|}{Evaluation Metrics for Stroke Prediction Dataset - Train set} \\
        \hline
        Model & Accuracy & Precision & Recall & F1-Score \\
        \hline\hline
        Logistic Regression (Manual) & 0.957191780 & 0. & 0.0 & 0.0 \\
        \hline
        Logistic Regression (Scikit-Learn) & 0.957436399 & \textbf{1.0} & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Manual) & 0.957436399 & \textbf{1.0} & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Scikit-Learn) & \textbf{0.96232876} & 0.83333333 & \textbf{0.14367816} &  \textbf{0.24509803} \\
        \hline
        \hline
        \multicolumn{5}{|c|}{Evaluation Metrics for Stroke Prediction Dataset - Test set} \\
        \hline
        Model & Accuracy & Precision & Recall & F1-Score \\
        \hline\hline
        Logistic Regression (Manual) & 0.925636007 & 0.0 & 0.0 & 0.0 \\
        \hline
        Logistic Regression (Scikit-Learn) & 0.925636007 & 0.0 & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Manual) & \textbf{0.92661448} & \textbf{1.0} & 0.0 & 0.0 \\
        \hline
        Multi-Layered Perceptron (Scikit-Learn) & 0.92367906 & 0.2 & \textbf{0.01333333} & \textbf{0.025} \\
        \hline

        \caption{Evaluation Metrics for Logistic Regression and Multi-Layered Perceptron Models on Stroke Prediction Dataset}
        \label{tab:evaluation_metrics_stroke} \\
    \end{longtable}
\end{center}

% ------------------------------------------------------------------------------
\section{Conclusions}
In this document, we explored the application of machine learning algorithms to
two distinct prediction tasks: Salary Prediction and Stroke Prediction. We conducted
a comprehensive analysis of the datasets, including data exploration, feature
correlation analysis, and data preprocessing. We implemented two prominent machine
learning algorithms, Logistic Regression and Multi-Layered Perceptron (MLP), to
predict the outcomes of the two tasks. We evaluated the models using various
evaluation metrics, including accuracy, precision, recall, and F1-Score, to assess
their performance.

The results of the evaluation metrics indicate that the models achieved varying
levels of performance on the two prediction tasks. The Logistic Regression model
performed well on the Salary Prediction dataset, achieving high accuracy and F1-Score
values. The Multi-Layered Perceptron model also demonstrated strong performance on
the Salary Prediction dataset, achieving high accuracy and F1-Score values. However,
the models struggled to predict the positive class effectively
on the Stroke Prediction dataset due to the high class imbalance. The Logistic
Regression and Multi-Layered Perceptron models achieved similar results on the
Stroke Prediction dataset, with low recall and F1-Score values for the positive class.

Overall, it is evident that the Multi-Layered Perceptron (Scikit-Learn 
implementation) generally performs better than Logistic Regression across both 
datasets. Specifically:

\begin{itemize}
    \item \textit{Salary Prediction:} The MLP model achieves higher accuracy, 
    precision, recall, and F1-score on the training set compared to Logistic
     Regression. While its test set performance is slightly lower than its 
     training set performance, it still demonstrates a competitive F1-score.
    \item \textit{Stroke Prediction:} Despite the challenge posed by class 
    imbalance, the MLP model achieves a better balance between precision and 
    recall than Logistic Regression, which fails to predict the positive class 
    at all, resulting in an F1-score of 0.
\end{itemize}

Therefore, based on these metrics, the Multi-Layered Perceptron (Scikit-Learn 
implementation) is the better algorithm for both the Salary Prediction and 
Stroke Prediction tasks due to its superior performance across multiple 
evaluation criteria.

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://curs.upb.ro/2023/course/view.php?id=13749}{Moodle - Artifical Intelligence Course}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Machine_learning}{Wikipedia - Machine Learning}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Logistic_regression}{Wikipedia - Logistic Regression}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Multilayer_perceptron}{Wikipedia - Multilayer Perceptron}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Ridge_regression}{Wikipedia - Ridge Regression}
    \bibitem{}
    \href{https://www.semanticscholar.org/reader/2c9022fe0af15568a885e59d475ec8f95726e51b}{Metrics for Multi-Class Classification: An Overview}
    \bibitem{}
    \href{https://developers.google.com/machine-learning/crash-course/classification/accuracy}{Google Developers - Classification: Accuracy}
    \bibitem{}
    \href{https://ocw.mit.edu/courses/6-867-machine-learning-fall-2006/}{MIT OpenCourseWare - Machine Learning}
    \bibitem{}
    \href{https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html}{Pandas - DataFrame}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Imputation_(statistics)}{Wikipedia - Imputation (statistics)}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Correlation}{Wikipedia - Correlation}
    \bibitem{}
    \href{https://thinkingneuron.com/how-to-measure-the-correlation-between-two-categorical-variables-in-python/}{Thinking Neuron - How to Measure the Correlation Between Two Categorical Variables in Python}
    \bibitem{}
    \href{https://www.stratascratch.com/blog/chi-square-test-in-python-a-technical-guide/}{StrataScratch - Chi-Square Test in Python: A Technical Guide}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%27s\_V}{Wikipedia - Cramér's V}
    \bibitem{}
    \href{https://saturncloud.io/blog/how-to-detect-and-exclude-outliers-in-a-pandas-dataframe/}{Saturn Cloud - How to Detect and Exclude Outliers in a Pandas DataFrame}
    \bibitem{}
    \href{https://www.kaggle.com/discussions/questions-and-answers/159183}{Kaggle - When to standardize test and train data?}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/preprocessing.html#standardization-or-mean-removal-and-variance-scaling}{Scikit-Learn - Standardization or Mean Removal and Variance Scaling}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.preprocessing.OneHotEncoder.html}{Scikit-Learn - OneHotEncoder}
    \bibitem{}
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.preprocessing.LabelEncoder.html}{Scikit-Learn - LabelEncoder}
    \bibitem{}
    \href{https://towardsdatascience.com/understanding-confusion-matrix-a9ad42dcfd62}{Towards Data Science - Understanding Confusion Matrix}
